Engineering an infinite-range SU($n$) spin model with ultracold fermions in an optical lattice

### 2021-04-26 sun_phases/spin_model_paper/main.tex
updates to paper

\begin{abstract}
We study multilevel fermions in an optical lattice, with dynamics described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a multilevel spin model with all-to-all SU($n$)-symmetric couplings.
Additionally, Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with the Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, beverland2016realizing, cazalilla2014ultracold, taie2012su, hofrichter2016direct, cappellini2014direct, scazza2014observation, zhang2014spectroscopic, goban2018emergence, perlin2019effective} with exquisite degrees of microscopic control.
As a simple example, this symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes the experimental observations of bosonization and faster thermalization in multi-component Fermi gas with $SU(n)$ symmetry \cite{song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, two-orbital SU($n$) magnetism \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic}, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.
In the spirit of quantum simulation, further investigations in controlled settings will play an important role in understanding the consequence of SU($n$) symmetries for fundamental questions in physics, as well as their practical use in technological applications.
For example, SU($2$)-symmetric spin interactions can be harnessed to develop quantum sensors that surpass classical limits on measurement precision \cite{he2019engineering, perlin2020spin}.
The prospect of similarly exploiting more general SU($n$) symmetries to achieve a technological advantage is still an unexplored avenue of research with untapped potential.

In this work, we consider an experimentally relevant and theoretically tractable regime of the SU($n$) Hubbard model, highlighting differences and similarities with the more familiar case of SU($2$).
Working at ultracold temperatures and unit spatial filling (one atom per lattice site), we begin by mapping the SU($n$) Hubbard model onto a multilevel spin model with all-to-all SU($n$)-symmetric interactions in Section \ref{sec:spin_model}.
We consider the effect of spin-orbit coupling (SOC) in Section \ref{sec:SOC}, finding in particular that the weak-SOC limit generally gives rise to a (synthetic) inhomogeneous magnetic field, extending previously known results to $n>2$ \cite{wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
In Section \ref{sec:controls} we examine the use of control fields to address nuclear spins (and generate SOC), finding a simple three-laser driving scheme that allows for the preparation of interesting states with nontrivial intra-spin correlations when $n>2$.
Finally, we combine these ingredients to examine mean-field dynamical behaviors of the SU($n$) spin model in Section \ref{sec:mean_field}, finding that:
\begin{enumerate*}
\item long-time-averaged observables obey simple scaling relations with the spin dimension $n$, exhibiting (for spin-polarized initial states) dynamical ferromagnetic and dynamical paramegnetic phases, as previously seen for the case of $n=2$ \cite{smale2019observation, lewis-swan2021cavityqed}, and
\item for $n>2$ the long-time dynamics can be highly sensitive to the intra-spin coherences of the initial state.
\end{enumerate*}
We conclude and discuss future directions in Section \ref{sec:conclusions}.

@@

Here we review the realization of a collective SU($n$) spin model in a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.}, \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_k e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

@@

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU($2$) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \v\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v{\hat{A}},\v{\hat{B}}$, we define the inner product
\begin{align}
  \v{\hat{A}} \cdot \v{\hat{B}}
  \equiv \sum_{\mu,\nu} A_{\mu\nu}^\dag B_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \v\s_p\cdot\v\s_q
  = -\f{u}{2N}\v\S\cdot\v\S,
  &&
  \v\S \equiv \sum_q \v\s_q.
  \label{eq:H_int}
\end{align}
Here $\v\S$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU($2$) \cite{he2019engineering}, with $\frac12\v\S\cdot\v\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality after modding out identity terms).

@@

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v\s_p\cdot\v\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v\s_p\cdot\v\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v\s_p\cdot\v\s_q$ is analogous to the enforcement of SU($2$) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\v\s_p\cdot\v\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

@@

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a probability distribution\footnote{Strictly speaking, in order to make $Q_\psi$ (and $Q_{\hat\rho}$) a probability distribution, one has to divide it by a normalization factor $\frac{4\pi}{d}$.
We omit this normalization factor for convenience.} $Q_\psi$ on the sphere $\SS^2$.
The value $Q_\psi\p{\v v}$ at a point $\v v\in\SS^2$ equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized along $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\equiv\bk{\v v|\hat\rho|\v v}$\footnote{The dimension $n$ is reflected in $Q_{\hat\rho}$ by the fact that it can be decomposed into spherical harmonics $Y_{\ell m}$ with degree $\ell<n$ (that is, $Q_{\hat\rho}$ is ``low resolution'' or ``band-limited'' at degree $n$) \cite{perlin2020qudit}, as well as in the normalization $\int_{\v v\in\SS^2}\dd^2\v v\,Q_{\hat\rho}\p{\v v}=\frac{4\pi}{d}$.}.
Closely related spherical representations of multilevel spin states (and operators) are discussed in Ref.~\cite{li2013weylwignermoyal}.

@@

In fact, $Q_{\hat\rho}$ is essentially the Husimi-$Q$ function commonly used (e.g.~in the spin squeezing community \cite{ma2011quantum}) to represent collective states of spin-$\frac{1}{2}$ particles in the Dicke manifold.
This correspondence is made precise by identifying the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac{1}{2}$ particles.
The Hilbert space of a 10-level (spin-$\frac92$) nucleus of a ${}^{87}$Sr atom, for example, can be thought of as the Dicke manifold of 9 spin-$\frac{1}{2}$ particles (``excess'' protons and neutrons).
We will freely use this correspondence to import standard results and constructions for the Dicke manifold (e.g.~the construction of spin operators $\s_\x,\s_\y,\s_\z$) into the context of a multilevel spin.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Spin-orbit coupling and gauge transformation}
\label{sec:SOC}

We wish to study the dynamical behavior of experimentally realizable SU($n$) spin models.
In order to ensure validity of the $\v\S\cdot\v\S$ spin Hamiltonian in Eq.~\eqref{eq:H_int}, we consider initial preparation of a spin-polarized state\footnote{We remind the reader that validity of the spin model requires that each momentum mode is at most singly-occupied, which is guaranteed for a spin-polarized state due to Pauli blocking.}, which has the added benefit of being simple to prepare experimentally.
However, a spin polarized state is an eigenstate of the $\v\S\cdot\v\S$ spin Hamiltonian, so we need additional ingredients for nontrivial behavior.
Here, we consider the addition of spin-orbit coupling (SOC) induced by external driving fields.

Before discussing SOC for $n$-level fermions, we first consider the well-studied case of two-level SOC with a one-dimensional lattice \cite{wall2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
In this case, SOC is induced by an external driving field that imprints a phase $e^{-\i\phi j}$ on lattice site $j$, or equivalently imparts a momentum kick $q\to q+\phi$, upon the absorption of a photon\footnote{In order for the drive Hamiltonian $\H_{\t{drive}}^{(\phi)}$ to be well-defined, $\phi$ should be commensurate with the lattice, with e.g.~$\phi\in2\pi/L\times\ZZ$ on a one-dimensional lattice of $L$ sites.}:
\begin{align}
  \H_{\t{drive}}^{(\phi)}
  = \f{\Omega}{2} \sum_q \c_{q+\phi,\up}^\dag \c_{q,\dn} + \t{h.c.}.
  \label{eq:drive_2}
\end{align}
Identifying a numerical spin index $\mu=+\frac12$ ($-\frac12$) with the state $\up$ ($\dn$), this drive Hamiltonian can be diagonalized in its momentum index $q$ by the gauge transformation $\c_{q\mu}^\dag\to \c_{q-\mu\phi,\mu}^\dag$ (equivalently $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$), which takes
\begin{align}
  \H_{\t{drive}}^{(\phi)} \to \H_{\t{drive}} \equiv \Omega \S_\x,
  &&
  \S_\x \equiv \sum_q \s_{\x,q},
  \label{eq:drive_trans}
\end{align}
where $\S_\x$ is a collective spin-x operator, and $\s_{\x,q}=\frac12 \c_{q,\up}^\dag \c_{q,\dn} + \t{h.c.}$ for two-level spins.

The two-level SOC drive in Eq.~\eqref{eq:drive_2} is typically implemented by coupling two electronic states of an atom with an external laser \cite{wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
Though the coupling between optical laser fields and nuclear spins is generally too weak to drive nuclear spin transitions directly, it will nonetheless be useful to consider the $n$-level generalization of the two-level drive.
This generalization will capture the important features of SOC for $n$-level spins, while avoiding complications that we defer to Section \ref{sec:controls}.

The spin-$s$ generalization of Eq.~\eqref{eq:drive_2} is
\begin{align}
  \H_{\t{drive}}^{(\phi)} = \Omega_0 \sum_{q,\mu} \bk{s,\mu;1,1|s,\mu+1}
  \c_{q+\phi,\mu+1}^\dag \c_{q\mu} + \t{h.c.},
\end{align}
where $\mu\in\set{s,s-1,\cdots,-s}$, and $\bk{s,\mu;1,1|s,\mu+1}$ is a Clebsch-Gordan coefficient.
The gauge gauge transformation $\c_{q\mu}^\dag\to \c_{q-\mu\phi,\mu}^\dag$ again takes $\H_{\t{drive}}^{(\phi)}\to \Omega \S_\x$ as in Eq.~\eqref{eq:drive_trans}, where now
\begin{align}
  \Omega = \f{\Omega_0}{\xi},
  &&
  \xi \equiv -\sqrt{\f{s(s+1)}{2}},
\end{align}
and
\begin{align}
  \s_{\x,q} \equiv \xi \sum_\mu \bk{s,\mu;1,1|s,\mu+1} \c_{q,\mu+1}^\dag \c_{q\mu} + \t{h.c.}
  \label{eq:spin_x_op}
\end{align}
is an $n$-level spin-x operator for spin $q$, with which $e^{-\i\theta \s_{\x,q}}$ rotates the state of spin $q$ about the x axis by an angle $\theta$ (literally rotating the sphere in Figure \ref{fig:spin_dist}).
Despite complicating details, the change from two-level to $n$-level spins thus results in a drive Hamiltonian with an identical physical interpretation: after moving into an appropriate ``gauged frame'', this drive simply rotates all spins about the x axis at a rate $\Omega$.

\begin{figure}
\centering
\includegraphics{soc_panels.pdf}
\caption{
Spin-orbit coupling for 2-level ({\bf a},{\bf c}) and 4-level ({\bf b},{\bf d}) spins.
Colors indicate different spin projections $\mu$.
In the ``lab frame'' ({\bf a},{\bf b}), kinetic energy is insensitive to spin, but the drive imparts a momentum kick $q\to q+\phi$ into any atom that absorbs a photon.
Changing to the ``gauged frame'' ({\bf c},{\bf d}), essentially by shifting the momentum label $q$ for each spin state $\mu$, makes the drive diagonal in the momentum index, but comes at the cost of making kinetic energy spin-dependent.
}
\label{fig:soc_panels}
\end{figure}

>>>>

\begin{abstract}
We study multilevel fermions in an optical lattice, with dynamics described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a multilevel spin model with all-to-all SU($n$)-symmetric couplings.
Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms with exquisite degrees of microscopic control.
This symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes the experimental observations of bosonization and faster thermalization in multi-component Fermi gas with SU($n$) symmetry \cite{hazzard2012hightemperature, messio2012entropy, bonnes2012adiabatic, stellmer2013degenerate, yip2014theory, pagano2014onedimensional, choudhury2020collective, song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, two-orbital SU($n$) magnetism \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic, beverland2016realizing}, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.

In the spirit of quantum simulation, further investigations in controlled settings will play an important role in understanding the consequence of SU($n$) symmetries for fundamental questions in physics, as well as their practical use in technological applications.
For example, SU($2$)-symmetric spin interactions can be harnessed to develop quantum sensors that surpass classical limits on measurement precision \cite{he2019engineering, perlin2020spin}.
The prospect of similarly exploiting more general SU($n$) symmetries to achieve a technological advantage is still an unexplored avenue of research with untapped potential.

In this work, we consider an experimentally relevant and theoretically tractable regime of the SU($n$) Hubbard model, highlighting differences and similarities with the more familiar case of SU($2$).
Working at ultracold temperatures and unit spatial filling (one atom per lattice site), we begin by mapping the SU($n$) Hubbard model onto a multilevel spin model with all-to-all SU($n$)-symmetric interactions in Section \ref{sec:spin_model}.
In Section \ref{sec:controls} we consider the use of control fields to address nuclear spins, finding a simple three-laser driving scheme that allows for the preparation of interesting states with nontrivial intra-spin correlations when $n>2$.
We consider the effect of spin-orbit coupling (SOC) induced by control fields in Section \ref{sec:SOC}, finding in particular that the weak-SOC limit generally gives rise to a (synthetic) inhomogeneous magnetic field, extending previously known results to $n>2$ \cite{mancini2015observation, wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
Finally, we combine these ingredients to examine mean-field dynamical behaviors of the SU($n$) spin model in Section \ref{sec:mean_field}, finding that:
\begin{enumerate*}
\item long-time-averaged observables obey simple scaling relations with the spin dimension $n$, exhibiting (for spin-polarized initial states) dynamical ferromagnetic and dynamical paramegnetic phases, as previously seen for the case of $n=2$ \cite{smale2019observation, lewis-swan2021cavityqed}, and
\item for $n>2$ the long-time dynamics can be highly sensitive to the intra-spin coherences of the initial state.
\end{enumerate*}
We conclude and discuss future directions in Section \ref{sec:conclusions}.

@@

Here we derive a collective SU($n$) spin model for a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.}, \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_k e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

@@

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU($2$) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \v\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v{\hat{A}},\v{\hat{B}}$, we define the inner product
\begin{align}
  \v{\hat{A}} \cdot \v{\hat{B}}
  \equiv \sum_{\mu,\nu} \hat A_{\mu\nu}^\dag \hat B_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \v\s_p\cdot\v\s_q
  = -\f{u}{2N}\v\S\cdot\v\S,
  \label{eq:H_int}
\end{align}
where $\v\S\equiv\sum_q\v\s_q$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU($2$) \cite{he2019engineering}, with $\frac12\v\S\cdot\v\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality up to identity terms).

@@

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v\s_p\cdot\v\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v\s_p\cdot\v\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v\s_p\cdot\v\s_q$ is analogous to the enforcement of SU($2$) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $-\v\s_p\cdot\v\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

@@

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a probability distribution\footnote{Strictly speaking, in order to make $Q_\psi$ (and $Q_{\hat\rho}$) a probability distribution, one has to divide it by a normalization factor $\frac{4\pi}{d}$.
We omit this normalization factor for convenience.} $Q_\psi$ on the sphere $\SS^2$.
The value $Q_\psi\p{\v v}$ at a point $\v v\in\SS^2$ equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized along $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\equiv\bk{\v v|\hat\rho|\v v}$\footnote{The dimension $n$ is reflected in $Q_{\hat\rho}$ by the fact that it can be decomposed into spherical harmonics $Y_{\ell m}$ with degree $\ell<n$ (that is, $Q_{\hat\rho}$ is ``low resolution'' or ``band-limited'' at degree $n$) \cite{perlin2020qudit}, as well as in the normalization $\int_{\v v\in\SS^2}\dd\v v\,Q_{\hat\rho}\p{\v v}=\frac{4\pi}{d}$.}.
Closely related spherical representations of multilevel spin states (and operators) are discussed in Ref.~\cite{li2013weylwignermoyal}.

@@

In fact, $Q_{\hat\rho}$ is essentially the Husimi-$Q$ function commonly used (e.g.~in the spin squeezing community \cite{ma2011quantum}) to represent collective states of spin-$\frac{1}{2}$ particles in the Dicke manifold.
This correspondence is made precise by identifying the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac{1}{2}$ particles.
The Hilbert space of a 10-level (spin-$\frac92$) nucleus of a ${}^{87}$Sr atom, for example, can be thought of as the Dicke manifold of 9 spin-$\frac{1}{2}$ particles (``excess'' protons and neutrons).
We will freely use this correspondence to import standard results and constructions for the Dicke manifold (e.g.~the spin operators $\s_\x,\s_\y,\s_\z$) into the context of a multilevel spin.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{External control fields}
\label{sec:controls}

We now consider the addition of external control fields to address atoms' internal spin states, which will determine the observables we can access and initial states we can prepare.
Specifically, we consider off-resonantly addressing an electronic $\ket\g\to\ket\e$ transition of the atoms, and then perturbatively eliminating electronic $\ket\e$ excitations to arrive at an effective ground-state Hamiltonian addressing nuclear spins.
For simplicity, we will assume that the total spin $s$ of the ground- and excited-state (hyperfine) manifolds are the same, as e.g.~with the ${^1}\t{S}_0\to{^3}\t{P}_0$ transition of alkaline-earth-like atoms (AEAs).
However, the results of this section (namely the general form of effective nuclear spin Hamiltonians, as well as the corresponding set of accessible observables and initial states) are the same for transitions that take $s\to s\pm1$, so in practice one is free to address the hyperfine manifolds of the ${^1}\t{S}_0\to{^3}\t{P}_1$ transition of AEAs.

We assume that all driving lasers are detuned by $\Delta$ below an electronic transition.
Each laser is indexed by an axis of propagation, $\v v$, and a drive amplitude $\Omega_{\v v,+}$ ($\Omega_{\v v,-}$) for right-circularly (left-circularly) polarized light.
An electronic $\ket\g\to\ket\e$ excitation is thus accompanied by a nuclear spin transition $\mu_{\v v}\to\mu_{\v v}\pm1$, where $\mu_{\v v}$ is the projection of nuclear spin onto the axis $\v v$.
Furthermore, a laser pointing along $\v v$ will generally imprint a phase $e^{-\i\kappa\v v\cdot\v\ell j}$ on lattice site $j$, where $\kappa$ is the wavenumber of the laser (in units with lattice spacing $a=1$) and $\v\ell$ is a fixed unit vector parallel to the lattice.
On a one-dimensional lattice with transverse confinement, the multi-laser drive Hamiltonian is then
\begin{align}
  \H_{\t{drive}}^{\t{full}}
  = \sum_{j,\v v,\sigma} \Omega_{\v v\sigma}
  \p{e^{-\i\kappa\v v\cdot\v\ell j} \s_{\v v\sigma j}\otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat N_\e,
  \label{eq:H_drive_full}
\end{align}
where $\sigma\in\set{+1,-1}$ indexes a polarization of light; $\s_{\v v\sigma j}$ is an $n$-level spin-raising ($\sigma=+1$) or spin-lowering ($\sigma=-1$) operator along axis $\v v$ for an atom on lattice site $j$ (identical to that for $n-1$ spin-$\frac12$ particles in the Dicke manifold); $\ket\g_j$ and $\ket\e_j$ respectively denote the ground and excited electronic states of atom $j$; and $\hat N_\e=\1\otimes\sum_j\op\e_j$ counts the number of excited atoms (with $\1$ the identity operator on all spin degrees of freedom).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{3LD.pdf}
\caption{
Sketch of the three-laser drive used to address nuclear spins on a one-dimensional lattice.
Two counter-propagating lasers with right-circular polarization and amplitudes $\Omega_\pm$ point at an angle $\theta$ to the lattice axis.
A third, linearly polarized laser with amplitude $\Omega_0$ points in a direction orthogonal to both the lattice and the other driving lasers.
Absorbing a photon from the laser with amplitude $\Omega_m$ induces a transition $(\g,\mu)\to(\e,\mu+m)$ for the (electronic, nuclear spin) state of an atom, where nuclear spin is quantized along the z axis.
}
\label{fig:3LD}
\end{figure}
